---
title: "A3: Incarceration"
author: "Jonathan Cao"
output: html_document
---
```{r setup, include=FALSE}
# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Install required packages
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
}
if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr")
}
if (!requireNamespace("leaflet", quietly = TRUE)) {
  install.packages("leaflet")
}
if (!requireNamespace("tidyr", quietly = TRUE)) {
  install.packages("tidyr")
}
if (!requireNamespace("maps", quietly = TRUE)) {
  install.packages("maps")
}
if (!requireNamespace("viridis", quietly = TRUE)) {
  install.packages("viridis")
}

# Load required packages
library(ggplot2)
library(dplyr)
library(leaflet)
library(tidyr)
library(maps)
library(viridis)
```

### Introduction

Incarceration in the United States is increasing at unprecedented rates, and our country has the highest incarceration rate in the world. Racial disparities are a crucial explanation for this problem, notably among Black Americans, who are incarcerated at a considerably higher rate than any other race while being a minority in the country. I will concentrate on the prison and jail populations throughout time. Identifying the elements that influence race and gender in incarceration. The variables considered include incarceration rates per 100,000 persons for various racial groups and genders, as well as total incarceration rates for various states and counties as reported by the Vera Institute. These factors will assist demonstrate systematic racism and gender inequities in the United States criminal justice system.

### Summary Information

Using the dataset provided by the Vera Institute, I calculated the following key values with this code:

```{r summary, echo=FALSE, message=FALSE, warning=FALSE}
# Load the dataset
url <- "https://raw.githubusercontent.com/melaniewalsh/Neat-Datasets/main/us-prison-jail-rates.csv"
data <- read.csv(url)

# Find the most recent year in the data
recent_year <- max(data$year, na.rm = TRUE)

# Filter the data for the most recent year
recent_data <- data %>% filter(year == recent_year)

# Calculate the average incarceration rate for the most recent year
average_incarceration_rate <- mean(recent_data$total_jail_pop_rate, na.rm = TRUE)

# Find the highest and lowest incarceration rates
highest_incarceration_rate <- max(recent_data$total_jail_pop_rate, na.rm = TRUE)
lowest_incarceration_rate <- min(recent_data$total_jail_pop_rate, na.rm = TRUE)
county_highest_rate <- recent_data$county_name[which.max(recent_data$total_jail_pop_rate)]
county_lowest_rate <- recent_data$county_name[which.min(recent_data$total_jail_pop_rate)]

# Calculate the change in incarceration rate for Black Americans over the last N years
start_year <- 1990
end_year <- recent_year
black_incarceration_start <- mean(data$black_jail_pop_rate[data$year == start_year], na.rm = TRUE)
black_incarceration_end <- mean(data$black_jail_pop_rate[data$year == end_year], na.rm = TRUE)
black_incarceration_change <- ((black_incarceration_end - black_incarceration_start) / black_incarceration_start) * 100

# Calculate the gender disparity in the most recent year
male_incarceration_rate <- mean(recent_data$male_jail_pop_rate, na.rm = TRUE)
female_incarceration_rate <- mean(recent_data$female_jail_pop_rate, na.rm = TRUE)
gender_disparity <- male_incarceration_rate / female_incarceration_rate
```

- The average incarceration rate across all counties in the most recent year available is `r format(round(average_incarceration_rate, 2), nsmall = 2)` per 100,000 people.
- The highest incarceration rate is found in `r county_highest_rate`, with a rate of `r format(round(highest_incarceration_rate, 2), nsmall = 2)` per 100,000 people.
- The lowest incarceration rate is found in `r county_lowest_rate`, with a rate of `r format(round(lowest_incarceration_rate, 2), nsmall = 2)` per 100,000 people.
- Over the last `r end_year - start_year` years, the incarceration rate for Black Americans has changed by `r format(round(black_incarceration_change, 2), nsmall = 2)`%.
- The gender disparity is evident, with men being incarcerated at a rate `r format(round(gender_disparity, 2), nsmall = 2)` times higher than women.

The variables demonstrate how Black American incarceration has actually decreased from 1990 to 2018. However, it does not remove the fact that Black Americans are still incarcerated at a much higher rate, and how there must be more targeted reforms, seeing how much more men are incarcerated as well.


### The Dataset

- The data for this analysis were gathered by the Vera Institute of Justice, a non-profit organization dedicated to criminal justice reform. 
- The U.S. Department of Justice Bureau of Justice Statistics (BJS) collected the data, which was augmented by data from state correctional facilities where federal data was unavailable. I used Rmarkdown code to generate the data, which simplified the required variables and calculations shown above.
- The data was gathered to provide a thorough overview of incarceration rates in the United States, including the overall prison population, jail population, and demographic split by race and gender.
- The dataset has around 160,000 rows.
- The dataset contains 20 columns and includes characteristics such as the year, state, county, total prison population, jail population, demographic breakdown, and more variables including overall jail, prison, and incarceration rates. When working with this data, ethical considerations must be made, especially when assessing the data's race and gender discrepancies in the United States' justice systems. It is critical to examine power dynamics and the possibility of perpetuating prejudices or biases in data analysis.

The data is gathered by the United States Department of Justice Bureau of Justice Statistics and augmented with information from state correctional facilities when federal data is unavailable. This imposes some constraints, as it implies that the data is collected by the government, which may introduce biases that they may conceal from the public. For instance, the data may not be collected consistently across all states, or there may be inconsistencies in how the data is obtained or published, making the country appear more "just" than it actually is. Furthermore, with so many factors at play and disparities, it is unavoidable that the data may contain mistakes and inconsistencies, which may impair the outcomes of the research. The data may not include every component of the criminal justice system, such as the influence of policing techniques or sentencing regulations. These issues are pervasive in our nation and may distort the results of such datasets because they contribute to the incarceration of victims. This may restrict the breadth of the analysis and preclude a thorough comprehension of the issues at hand. The statistics may also fail to capture the entire scope of racial disparities in the criminal justice system since it excludes data on other marginalized groups and gender disparities. This could hinder our capacity to identify and address the underlying causes of these differences, as well as develop effective solutions.

### Trends Over Time Chart

The following chart shows the trend of incarceration rates over time for Black and White Americans. It illustrates the significant racial disparities in incarceration rates and how these have evolved over the years, whether for the better or worse.

```{r race trends, echo = FALSE}
# Data processing
data_filtered <- data %>%
  filter(year >= 1990, !is.na(black_jail_pop_rate), !is.na(white_jail_pop_rate)) %>%
  group_by(year) %>%
  summarize(black_jail_pop_rate = mean(black_jail_pop_rate, na.rm = TRUE),
            white_jail_pop_rate = mean(white_jail_pop_rate, na.rm = TRUE)) %>%
  pivot_longer(cols = c(black_jail_pop_rate, white_jail_pop_rate),
               names_to = "Race",
               values_to = "IncarcerationRate")

# Create the trend chart
ggplot(data_filtered, aes(x=year, y=IncarcerationRate, color=Race)) +
  geom_line() +
  labs(title="Incarceration Rates Over Time by Race", x="Year", y="Incarceration Rate per 100,000 People") +
  theme_minimal() +
  scale_color_manual(values=c("black_jail_pop_rate"="black", "white_jail_pop_rate"="blue")) +
  theme(legend.title = element_blank())
```

- The chart clearly shows the Incarceration Rates per 100,000 people
- The chart shows the trend of incarceration rates over time for Black and White Americans
- Despite the incarceration rates for black Americans decreasing over time, it is still significantly higher than that of white Americans

### Variable Comparison Chart

This chart shows the gender disparities within incarceration rates, for males and females across different states.

```{r gender chart, echo = FALSE}
# Load the dataset from the URL
url <- "https://raw.githubusercontent.com/melaniewalsh/Neat-Datasets/main/us-prison-jail-rates.csv"
data <- read.csv(url)

# Data processing
data_comparison <- data %>%
  filter(year == max(year, na.rm = TRUE), !is.na(female_jail_pop_rate), !is.na(male_jail_pop_rate)) %>%
  group_by(state) %>%
  summarize(female_jail_pop_rate = mean(female_jail_pop_rate, na.rm = TRUE),
            male_jail_pop_rate = mean(male_jail_pop_rate, na.rm = TRUE)) %>%
  pivot_longer(cols = c(female_jail_pop_rate, male_jail_pop_rate), 
               names_to = "Gender", 
               values_to = "IncarcerationRate")

# Create a scatter plot comparing male and female incarceration rates by state
ggplot(data_comparison, aes(x=state, y=IncarcerationRate, color=Gender)) +
  geom_point(alpha=0.7) +
  labs(title="Comparison of Male and Female Incarceration Rates by State", 
       x="State", 
       y="Incarceration Rate per 100,000 People", 
       color="Gender") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_color_manual(values=c("female_jail_pop_rate"="pink", "male_jail_pop_rate"="blue"))
```

- The chart shows the Incarceration Rates per 100,000 people.
- The chart clearly identifies that much more males are incarcerated than females.

### Map

This chart shows a geographical distribution of incarceration rates across the United States by visualizing a map of the U.S. with colorcoded regions by incarceration rates per 100,000 people.

```{r map, echo=FALSE, message=FALSE, warning=FALSE}
# Load required packages
library(dplyr)
library(ggplot2)
library(maps)
library(viridis)

# Load the dataset from the URL
url <- "https://raw.githubusercontent.com/melaniewalsh/Neat-Datasets/main/us-prison-jail-rates-1990.csv"
geo_data <- read.csv(url)

# Process data for mapping
geo_data_filtered <- geo_data %>%
  filter(year == max(year, na.rm = TRUE), !is.na(total_jail_pop_rate)) %>%
  group_by(state) %>%
  summarize(total_jail_pop_rate = mean(total_jail_pop_rate, na.rm = TRUE))

# Ensure that the state names match those in the spatial data
geo_data_filtered$state <- tolower(state.name[match(geo_data_filtered$state, state.abb)])

# Create the map
states <- map_data("state")
geo_map <- merge(states, geo_data_filtered, by.x = "region", by.y = "state")

# Plot the map
ggplot(data = geo_map, aes(x = long, y = lat, group = group, fill = total_jail_pop_rate)) +
  geom_polygon(color = "white") +
  labs(title = "Geographical Distribution of Incarceration Rates",
       fill = "Incarceration Rate per 100,000 People") +
  theme_minimal() +
  scale_fill_viridis_c(option = "plasma")
```

- There are noticable regional variations in incarceration rates across the United States. States in the Northeast, Midwest, and West Coast, seem to have lower incarceration rates. States in the South and Southwest tend to have much higher incarceration rates. This may be a reflection of higher crime rates, or issues within the state's government policies that cause more discrepancies for incarcerations.